---
title: "Join wood densities into wide-to-long dataset"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
execute:
  echo: true
  warning: true
  message: true
params:
  # Put this .qmd in the same folder as your two files, or override below at render time.
  main_csv: "wide-to-long_data_COMBINED_with_parsed_flake_AND_fuelwood.csv"
  wood_xlsx: "wood densities.xlsx"
  outdir: "."
---

```{r}
#| label: setup
#| include: true
library(tidyverse)
library(readr)
library(readxl)
library(janitor)
library(glue)
library(fs)

# Create output folder
dir_create(params$outdir)

# Utility: squish whitespace and lower-case for robust joins
normalize_key <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_replace_all("\\s+", " ") %>%
    stringr::str_trim() %>%
    tolower()
}

# Utility: resolve a file path (try given path, else search by basename)
resolve_file <- function(path) {
  if (fs::file_exists(path)) return(path)
  bn <- basename(path)
  found <- tryCatch(
    fs::dir_ls(path=".", recurse=TRUE, type="file", regexp=paste0(bn,"$")),
    error=function(e) character(0)
  )
  if (length(found) > 0) return(found[[1]])
  stop(glue("File not found: {path}. Working dir: {getwd()}"))
}
```

## 1) Load both datasets

```{r}
#| label: load
main_path <- resolve_file(params$main_csv)
wood_path <- resolve_file(params$wood_xlsx)

message(glue("Main CSV:  {main_path}"))
message(glue("Wood XLSX: {wood_path}"))

main_raw <- readr::read_csv(main_path, show_col_types = FALSE)
wood_raw <- readxl::read_excel(wood_path)

# Clean column names to snake_case for safety
main <- janitor::clean_names(main_raw)
wood <- janitor::clean_names(wood_raw)

cat("Main columns:", paste(names(main), collapse=", "), "\n")
cat("Wood columns:", paste(names(wood), collapse=", "), "\n")
```

## 2) Prepare keys and density field

```{r}
#| label: prep-keys
# Ensure there is a 'fuelwood' column in both
if (!"fuelwood" %in% names(main)) stop("The main CSV must contain a 'fuelwood' column.")
if (!"fuelwood" %in% names(wood)) {
  # try to map likely variants
  poss <- grep("^fuel[_ ]?wood$|^wood$", names(wood), ignore.case = TRUE, value = TRUE)
  if (length(poss) == 0) stop("The wood densities file needs a 'fuelwood' column (or a close variant).")
  if (length(poss) > 0) {
    message(glue("Mapping wood densities column '{poss[1]}' -> 'fuelwood'"))
    wood <- wood %>% rename(fuelwood = all_of(poss[1]))
  }
}

# Ensure there is a numeric wood density column; prefer 'wood_density'
density_col <- NULL
if ("wood_density" %in% names(wood)) density_col <- "wood_density"
if (is.null(density_col)) {
  # look for candidates like density, dens_g_cm3, etc.
  cand <- grep("wood.*dens|^density$|dens|g.?/?.?cm3|gcm3|gcc|specific_gravity", names(wood),
               ignore.case = TRUE, value = TRUE)
  if (length(cand) == 0) {
    stop("Could not find a wood density column in the wood densities file.")
  } else {
    density_col <- cand[1]
    message(glue("Using '{density_col}' as wood density and renaming to 'wood_density'"))
    wood <- wood %>% rename(wood_density = all_of(density_col))
  }

# Ensure there is a wood density group column named 'wood_density_group'
if (!"wood_density_group" %in% names(wood)) {
  cand_grp <- grep("group", names(wood), ignore.case = TRUE, value = TRUE)
  if (length(cand_grp) > 0) {
    message(glue("Using '{cand_grp[1]}' as wood density group and renaming to 'wood_density_group'"))
    wood <- wood %>% rename(wood_density_group = all_of(cand_grp[1]))
  } else {
    message("No 'wood_density_group' column found in wood densities file; creating as NA.")
    wood <- wood %>% mutate(wood_density_group = NA_character_)
  }
}
}

# Normalize the join key for both frames
main <- main %>% mutate(fuelwood_key = normalize_key(fuelwood))
wood <- wood %>% mutate(fuelwood_key = normalize_key(fuelwood))

# Check for missing keys
main_missing <- sum(is.na(main$fuelwood_key) | main$fuelwood_key == "")
wood_missing <- sum(is.na(wood$fuelwood_key) | wood$fuelwood_key == "")
message(glue("Empty/NA fuelwood keys — main: {main_missing}, wood: {wood_missing}"))
```

## 3) Diagnose duplicates in wood table and de-duplicate safely

```{r}
#| label: dedup-wood
wood_dupes <- wood %>%
  count(fuelwood_key, name = "n_rows") %>%
  filter(n_rows > 1)

if (nrow(wood_dupes) > 0) {
  message(glue("Found {nrow(wood_dupes)} duplicated fuelwood keys in wood densities; aggregating by mean density."))
  # Save the list for review
  write_csv(wood_dupes, file.path(params$outdir, "wood_duplicates.csv"))
}

# Aggregate wood densities by mean per key to avoid row multiplication on join
wood_agg <- wood %>%
  group_by(fuelwood_key) %>%
  summarise(
    wood_density = suppressWarnings(mean(as.numeric(wood_density), na.rm = TRUE)),
    wood_density_group = dplyr::first(na.omit(wood_density_group)),
    n_source_rows = dplyr::n(),
    .groups = "drop"
  )
```

# Check for inconsistent grouping labels within the same fuelwood_key
grp_check <- wood %>%
  group_by(fuelwood_key) %>%
  summarise(n_groups = n_distinct(wood_density_group, na.rm = TRUE), .groups = "drop") %>%
  filter(n_groups > 1)
if (nrow(grp_check) > 0) {
  warning("Some fuelwood_key values have multiple wood_density_group labels in the source; the join uses the first non-NA label per key.")
}




## 4) Left join and verify

```{r}
#| label: join-and-verify
n_before <- nrow(main)
merged <- main %>%
  left_join(wood_agg, by = "fuelwood_key")

n_after <- nrow(merged)
if (n_after != n_before) {
  warning(glue("Row count changed after join (before: {n_before}, after: {n_after})."))
} else {
  message(glue("Row count unchanged after join: {n_after} rows."))
}

# How many rows didn't find a wood density?
n_na_density <- sum(is.na(merged$wood_density))
pct_na <- round(100 * n_na_density / nrow(merged), 2)
message(glue("Rows with missing wood_density after join: {n_na_density} ({pct_na}%)."))

# Identify any fuelwood keys in main that failed to match
unmatched <- merged %>%
  filter(is.na(wood_density)) %>%
  distinct(fuelwood, fuelwood_key) %>%
  arrange(fuelwood)
if (nrow(unmatched) > 0) {
  write_csv(unmatched, file.path(params$outdir, "unmatched_fuelwood_in_main.csv"))
  message(glue("Wrote list of unmatched main fuelwood values to 'unmatched_fuelwood_in_main.csv'"))
}

# Keep original fuelwood, drop helper key if you like; retain diagnostic columns
merged_out <- merged %>% relocate(wood_density, wood_density_group, .after = fuelwood)

# Quick peek
merged_out %>% select(fuelwood, wood_density, wood_density_group) %>% head(12)
```

## 5) Save merged CSV

```{r}
#| label: save
out_csv <- file.path(params$outdir, "wide-to-long_with_wood_density.csv")
readr::write_csv(merged_out, out_csv)
message(glue("Saved merged CSV to: {out_csv}"))
```

## 6) Summary tables

```{r}
#| label: summary
# Summary by fuelwood
summary_by_fuelwood <- merged_out %>%
  group_by(fuelwood) %>%
  summarise(
    n = dplyr::n(),
    wood_density = unique(wood_density)[1],
    .groups = "drop"
  ) %>%
  arrange(fuelwood)

# Density completeness by method/scale/distance/heating
completeness <- merged_out %>%
  mutate(has_density = !is.na(wood_density)) %>%
  count(method, scale, distance, heating_condition, has_density) %>%
  tidyr::pivot_wider(names_from = has_density, values_from = n, values_fill = 0, names_prefix = "has_density_") %>%
  arrange(method, scale, distance, heating_condition)

# Print
summary_by_fuelwood %>% head(20) %>% knitr::kable(digits = 3, caption = "First 20 fuelwood entries with densities")
completeness %>% knitr::kable(digits = 0, caption = "Wood density completeness by strata")
```
